This is an example of article title

### 2019-06-02 paper/paper.tex
finished another review pass

Sorting has enjoyed an important place within computer programming topics for a long time. Apart from having great importance in practical applications, being very useful to organize data for fast retrieval, it also attracts attention as a theoretically interesting problem in itself~\cite{DBLP:books/lib/Knuth98a}. While sorting is a well-understood problem in general, with a few classic algorithms available, achieving optimal performance for specific problems and architectures may require a careful algorithm selection rather than just tuning parameters \cite[Part II, Introduction]{DBLP:books/daglib/0023376}.

The running time of a computer program is governed by many factors, starting with algorithmic complexity and the base clock speed of the processor. The performance observed in real computers has been increasingly depending in processor features that are not taken into account by the simplest models, though. These features include cache memory and paralellism at the processor and instruction levels. Cache memory has become popular in consumer processors in the past couple of decades, allowing programs to often attain higher performance than would be possible with a simpler memory architecture, as long as memory access patterns exhibit temporal and spatial locality~\cite[Chapter 3]{Drepper07whatevery}. Writing software that can better cope with parallelism has also become necessary to exploit the full potential offered by modern and future processors~\cite{wilson2018}.

The interest in sorting and the great relevance of cache memory and parallelism in modern computing naturally led to research on sorting algorithms that account for these factors. Processors offering SIMD (single instruction, multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s. One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}. The current availability of processors with 512 bit registers underscores the need of being able to write software that exploits vector operations, not to mention GPUs.

@@

\section{SIMD sorting techniques}
\label{sec:methods}
%
\chipsort offers four high-level functions that can be used in applications as replacements for other sorting functions such as \verb sort  in the Julia standard library. The {\tt chipsort!} function does not require any configuration and offers a great performance on a wide range of cases. The remaining functions are intended to arrays of different sizes, and require choosing some numeric parameters. The function \verb chipsort_large  offers a good overview of all the techniques implemented in the module, and this section is structured according to its stages. Some of these techniques might also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.

@@

\subsection{Sorting networks}
%
Sorting networks~\cite[Sec. 5.3.4]{DBLP:books/lib/Knuth98a} are graphs composed by comparator modules, units that take two numbers in, and output them in order: the smallest and the largest at specific outputs. Properly arranging the compatators allows the creation of a procedure that takes a vector of numbers in, and outputs them in sequence. The design of a sorting network can minimize the total number of comparisons, or also take into account that comparisons can be done in parallel, minimizing the number of steps to complete the process.

@@

\chipsort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}. The generated function \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} works by assigning each element of the sequence at each step to a variable, and calculating values from the next step according to the network specification. The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types. Figure~\ref{fig:sort-net-listing} shows one example of generated code.

@@

The {\tt sort\_net} generated function expects elements with {\tt max} and {\tt min} methods defined. It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works. In practice \chipsort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}. This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers, or spilling data to the stack if all registers are occupied.

\subsection{SIMD vectors transpose}
%
Consider a group of stacked SIMD vectors. Each column in this representation is termed a {\em lane}. The result from {\tt sort\_vecs} is that each lane contains a sorted sequence. This data must be transposed to produce sorted vectors. This is done in \chipsort through {\tt transpose\_vecs}, which is a generated function that supports rectangular matrices, but with dimensions constrained to powers of 2. The generated code consists mostly of calls to {\tt SIMD.shufflevec}. This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions contemplating each different case, while in \chipsort there is only a single Julia generated function that produces equivalent code.

Figure~\ref{fig:transpose-vecs} displays an example of 16 random integers loaded into SIMD registers as 4 vectors. After {\tt sort\_vecs} each lane contains a sorted sequence, and {\tt transpose\_vecs} turns the lanes into vectors (registers). The \verb chipsort_small!  function is essentially a sorting network followed by a transpose and a merge procedure.

@@

\subsection{Bitonic Merge Networks}
%
A bitonic sequence starts either increasing or decreasing and contains at most one change of direction. Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/bitonic-merge-network.jl#L12}{\tt bitonic\_merge}. Merging two sorted vectors requires first reversing one of them, and the result is stored into two new vectors of the same size containing the first and second halves of the total sorted sequence.

@@

\subsection{Vectorized Combsort}
%
One of the most peculiar aspects of \chipsort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm. Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{INCERPI198737,dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}. The array is swept multiple times, and pairs of values at a given distance from each other are compared and stored in order. The distance is reduced at each pass by a factor of $\frac{4}{3}$. This algorithm lends itself easily to vectorization, either by parallelizing the operations from a single array or executing the algorithm over independent arrays in parallel.

\chipsort contains two functions that utilize Combsort, employing parallelism in these two distinct forms. The first is {\tt chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the optimizing compiler. {\tt chipsort!} starts with Combsort until the interval size is 1, when it switches to Insertion sort. This is similar to what is often done in practical implementations of Quicksort, as in the Julia standard library, switching to Insertion sort when the array becomes too small. The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort is applied only once through the whole array.

@@

In both implementations switching to Insertion sort presents a better performance than sticking to Combsort until the end. It essentially means using Bubble sort once the interval becomes 1, therefore looking for an alternative seems like a good idea. Insertion sort should fare well because the array is now approximately sorted, and it should not have to move any out-of-order elements very far from where they are found.

Finishing a sorting method with Insertion sort is a common practice, present at least in Musser's Introsort~\cite{musser1997introspective}. The utilization of Insertion sort as a final stage to Combsort has been considered before~\cite{combwiki,INCERPI198737}, although the method does not seem to enjoy wide adoption. The main contribution of the present work may be to propose it again in the SIMD context, extending AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} and bringing experimental evidence of its good performance.

\subsection{In-place matrix transpose}
%
In-place matrix transposition can be attained by moving a number to its destination, then moving the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}. While carrying out this procedure is simple, it is not trivial to find out what are the cycles to tranpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[Sec. 1.3.3, Ex. 12]{DBLP:books/lib/Knuth97}.

\chipsort contains the generated function \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/transpose-vecs.jl#L19}{\tt transpose!} to perform in-place matrix transposition. The cycle seeds are computed at the time of code generation, and at runtime the function only needs to move the data and compute the indices detecting when each cycle finishes. The function is therefore completely generic, supporting any dimensions, but with a very simple runtime code, with the most complex part of the problem being tackled programatically during code generation.

@@

Given a set of sorted arrays, once the tree is initialized data is taken from it one vector at a time, producing the final sorted sequence. Each time a new vector is requested the tree is traversed looking for the smallest element between each of the node children, until one of the input arrays is reached. The data from the array is then loaded from memory and merged with the intermediate data, producing new updates moving trought the tree until the root is reached.

The intermediate data from the tree is kept by \chipsort in a contiguous array, treated as a heap. Assuming 1-based indexing, each node ancestor can be found simply by the Euclidean division of its index by 2. During operation it is assumed this data should fit some level of cache memory, as in the related work~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15}.

\section{Experiments}
\label{sec:experiments}
%
This section reports experiments with the \chipsort methods, as well as alternative ones. The term {\it method} can be understood as both the algorithms and their implementations in the Julia language. The specific methods tested are:
\begin{itemize}
\item {\tt chipsort!} Combsort finishing with Insertion sort, written as a serial program and vectorized by the compiler.
\item {\tt chipsort\_small!} in-register, based on sorting networks.
\item {\tt chipsort\_medium!} vectorized Combsort, plus additional steps, finishing with Insertion sort.
\item {\tt chipsort\_large} merge-sort with a multi-way merge tree, starting from chunks processed by {\tt chipsort\_medium!}.
\item {\tt insertion\_sort!} Insertion sort taken from the Julia library.
\item {\tt sort!} the Julia standard sort, Quicksort finishing with Insertion sort for small arrays (20 elements).
\end{itemize}

@@

\subsection{Medium and large arrays}
In our second experiment we compared many different methods on inputs of exponentially increasing sizes, from $2^6$ up to $2^{20}$. Figure~\ref{fig:bench-time} displays the resulting measurements. Although the curves look too concentrated in this visualization, it is possible to notice that {\tt chipsort!} consistently outperforms the Julia standard library {\tt sort!} by a factor of 80\% to 100\%, or up to twice as fast. We can also observe that Insertion sort soon becomes too inefficient.

\begin{figure}[htb]
\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-time.pdf}}
\caption{Running time of the different sorting methods studied.}
\label{fig:bench-time}
\end{figure}

Figure~\ref{fig:bench-curves} shows a different visualization from the same data in Figure~\ref{fig:bench-time}, with the times divided by the input size at each point. The vertical dashed lines indicate where the input array would be as big as each cache level. Here we can see more clearly how {\tt chipsort\_medium!} and {\tt chipsort\_large} slightly outperform {\tt chipsort!} in a few cases.

\begin{figure*}[thb]
\centerline{\includegraphics[width=0.7\linewidth]{fig/chipsort-bench-curves.pdf}}
\caption{Curves displaying the relative running time from each studied method. The ordinate represents the median of the measured running times divided by the input size at each test.}
\label{fig:bench-curves}
\end{figure*}

@@

\begin{figure}[htb]
\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-8k.pdf}}
\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-256k.pdf}}
\caption{Empirical CDF from different methods with 8k and 256k {\tt UInt32} values. Results from 10,000 measurements.}
\label{fig:bench-cdf}
\end{figure}

\subsection{Other languages}
A simple experiment was also carried out with different programming languages in the same machine, and with Julia in different machines to provide a richer context. Table~\ref{tab:bench-cxx} displays statistics collected from Julia, Python and C++ programs. For the case of Python the measurements obtained by calling the function through PyCall were better than using the native Python shell, therefore the best value was kept. Results obtained for C++ with Julia inter-operation using Cxx.jl were slightly worse than native, thus discarded. While Numpy was almost as fast as Julia, no reason could be found for the dismal performance from the C++ standard library.

\begin{table}[h]
\tabcolsep11pt
\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different platforms.}{
\begin{tabular}{|l|r|r|}\hline
Method & 8k & 1M\\\hline
\chipsort & 172.35 $\mu$s & 35.28 ms\\\hline
Julia standard & 345.46 $\mu$s & 68.39 ms\\\hline
Numpy via PyCall & 369.50 $\mu$s & 70.01 ms\\\hline
C++ {\tt std::qsort} & 788.59 $\mu$s & 147.37 ms\\\hline
C++ {\tt std::sort} & 1,207.23 $\mu$s & 224.71 ms\\\hline
\end{tabular}}
\label{tab:bench-cxx}
\end{table}

Table~\ref{tab:bench-512} contains the results for {\tt chipsort!} and the Julia standard {\tt sort!} in an Intel processor with AVX512 instructions, and an ARM processor with NEON instructions.
\begin{table}[h]
\tabcolsep11pt
\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different platforms.}{
\begin{tabular}{|l|r|r|}\hline
Method & 8k & 1M\\\hline
AXV512 \chipsort & 177.4 $\mu$s & 35.20 ms\\\hline
AXV512 Julia standard & 347.5 $\mu$s & 69.07 ms\\\hline
NEON \chipsort & 350.9 $\mu$s & 73.36 ms\\\hline
NEON Julia standard & 573.2 $\mu$s & 113.68 ms\\\hline
\end{tabular}}
\label{tab:bench-512}
\end{table}

No change in performance can be seen in the AVX512 machine, and on the ARM machine the speedup was smaller than before: only 60\%. The most notable fact in this experiment is the fact the same code was successfully employed to produce a vectorized sort on machines with different vector sizes and even instruction sets, always with beneficial effects.

\section{Conclusion}
\label{sec:conclusion}
%
We have presented \chipsort, a Julia module for SIMD and cache-aware sorting. The main sorting method utilized in \chipsort is Combsort finalized with Insertion sort. This method is based on previous proposals for SIMD sorting~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} and other practical sorting methods~\cite{INCERPI198737,musser1997introspective}. This method was implemented as a serial program vectorized by the Julia compiler and LLVM, resulting in a sorting function up to twice as fast as the Julia standard library sort. This performance holds for at least a million 32-bit integers in a conventional computer.

@@

While the main \chipsort method relies on compiler optimizations, other methods utilize explicit vectorization trough SIMD.jl. \chipsort makes extensive use of metaprogramming, more specifically Julia generated functions. This allows functions like sorting networks to be implemented in an abstract and generic way. The networks are represented as data structures, used by a higher-level program to assemble the final network code. This contrasts with other libraries where different networks are implemented straight as final functions. Julia also has the opportunity to perform many optimizations since \chipsort is implemented in pure Julia with abstract methods that can be specialized.

It is a testament to the quality of the Julia compiler infrastructure that the most successful method implemented in \chipsort has no explicit vectorization or intrinsics. It is just a simple, scalar implementation of Combsort that is optimized by the compiler to generate code adapted to different problem settings and to the hardware architecture at run-time. And in the cases where explicit vectorization can produce better code Julia still offers a suitable framework that allowed \chipsort to implement sophisticated techniques with concise code. This project can illustrate how well Julia implements traditional language features such as parametric types, and its metaprogramming features make it stand out from other languages even more.

Some future prospects for \chipsort are to better deal with floating-point comparisons; implement stable sorting of key-payload data; explore the limits of in-register processing, specially in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to produce optimized code for complex tasks such as multi-stage multi-threaded merge sorting of large arrays.

>>>>

Sorting has enjoyed an important place within computer programming topics for a long time. It has great importance to practical applications, being very useful to organize data for fast retrieval, and also attracts attention as a theoretically interesting problem in itself~\cite{DBLP:books/lib/Knuth98a}. While sorting is a well-understood problem in general, with a few classic algorithms available, achieving optimal performance for specific problems and architectures may require a careful algorithm selection rather than just tuning parameters \cite[Part II, Introduction]{DBLP:books/daglib/0023376}.

The running time of a computer program is governed by many factors, starting with algorithmic complexity and the base clock speed of the processor. The performance observed in real computers has been increasingly depending in processor features that are not taken into account by the simplest models, though. These features include cache memory and paralellism at the processor and instruction levels. Cache memory has become popular in consumer processors in the past couple of decades, and as long as memory access patterns exhibit temporal and spatial locality, it often allows programs to attain a higher performance than would be possible with a simpler memory architecture~\cite[Chapter 3]{Drepper07whatevery}. Writing software that can better cope with parallelism has also become necessary to exploit the full potential offered by modern and future processors~\cite{wilson2018}.

The interest in sorting and the great relevance of cache memory and parallelism to modern computing naturally led to research on sorting algorithms that account for these factors. Processors offering SIMD (single instruction, multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s. One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}. The current availability of processors with 512 bit registers underscores the need of being able to write software that exploits vector operations, what can also benefit GPU programming.

@@

\section{SIMD sorting techniques}
\label{sec:methods}
%
\chipsort offers four high-level functions that can be used in applications as replacements for other sorting functions such as {\tt sort} in the Julia standard library. The {\tt chipsort!} function does not require any configuration and offers a great performance on a wide range of cases. The remaining functions are suitable for arrays of specific size ranges, and require choosing some numeric parameters. The function \verb chipsort_large  offers a good overview of all the techniques implemented in the module, and this section is structured according to its stages. Some of these techniques may also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.

@@

\subsection{Sorting networks}
%
Sorting networks~\cite[Sec. 5.3.4]{DBLP:books/lib/Knuth98a} are graphs composed by comparator modules, units that take two numbers in, and output them in order: the smallest and the largest at specific outputs. By properly arranging the comparators we can create a procedure to takes a vector of elements and output them in sequence. The design of a sorting network can minimize the total number of comparisons, or if they can be done in parallel, minimize the number of steps to complete the process.

@@

\chipsort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}. The generated function \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} assigns each element of the sequence at each step to a variable, and the values from the next step are calculated according to the network specification. The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types. Figure~\ref{fig:sort-net-listing} shows one example of generated code.

@@

The {\tt sort\_net} generated function expects elements to which {\tt max} and {\tt min} are defined. It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works. In practice \chipsort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}. This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers, or spilling data to the stack if all registers are occupied.

\subsection{SIMD vectors transpose}
%
Consider a group of stacked SIMD vectors. Each column in this representation is termed a {\em lane}. The result from {\tt sort\_net} with vectors is that each lane contains a sorted sequence. This data must be transposed to produce sorted vectors. This is done in \chipsort through {\tt transpose\_vecs}, which is a generated function that supports rectangular matrices, but with dimensions constrained to powers of 2. The generated code consists mostly of calls to {\tt SIMD.shufflevec}. This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions contemplating each different case, while in \chipsort there is only a single Julia generated function that produces equivalent code.

Figure~\ref{fig:transpose-vecs} displays an example of 16 random integers loaded into SIMD registers as 4 vectors. After {\tt sort\_net} each lane contains a sorted sequence, and {\tt transpose\_vecs} turns the lanes into vectors (registers). The {\tt chipsort\_small!} function is essentially a sorting network followed by a transpose and a merge procedure.

@@

\subsection{Bitonic Merge Networks}
%
A bitonic sequence starts either increasing or decreasing and contains at most one change of direction. Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/bitonic-merge-network.jl#L12}{\tt bitonic\_merge}. Merging two sorted vectors requires first reversing one of them, and the result is stored in two vectors containing the first and second halves of the complete sorted sequence.

@@

\subsection{Vectorized Combsort}
%
One of the most peculiar aspects of \chipsort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm. Combsort essentially generalizes Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{INCERPI198737,dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}. The array is swept multiple times, ordering pairs of values at a given distance from each other. The distance is reduced each time by $\frac{4}{3}$. Parallelism is achieved either by parallelizing the operations from a single array or by processing independent arrays of the same size.

\chipsort contains two functions that utilize Combsort, employing these two distinct forms of parallelism. The first is {\tt chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the optimizing compiler. {\tt chipsort!} starts with Combsort until the interval size is 1, when it switches to Insertion sort.

@@

In both implementations switching to Insertion sort presents a better performance than sticking to Combsort until the end. Once the interval becomes 1 the algorithm essentially turns into Bubble sort, therefore looking for an alternative seems like a good idea. Insertion sort should fare well because the array is now approximately sorted, and any out-of-order elements found should be close to their destinations.

Finishing a sorting method with Insertion sort is a common practice, present at least in Musser's Introsort~\cite{musser1997introspective}. This is also akin to the classic method used in the Julia standard library, switching to Insertion sort when the array becomes too small. The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort must be applied once on the whole array.

The utilization of Insertion sort as a final stage to Combsort has been considered before~\cite{combwiki,INCERPI198737}, and Insertion sort was even employed by~\cite{DBLP:journals/pvldb/InoueT15} for dealing with partial keys. The main contribution of the present work may be to extend AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} by giving a larger role to Insertion sort and also relying more on automatic vectorization by the compiler.

\subsection{In-place matrix transpose}
%
In-place matrix transposition can be attained by moving a number to its destination, then moving the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}. While carrying out this procedure is simple, it is not trivial to find out what are the cycles to transpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[1.3.3-12]{DBLP:books/lib/Knuth97}.

\chipsort contains the generated function \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/transpose-vecs.jl#L19}{\tt transpose!} to perform in-place matrix transposition. The cycle seeds are computed at the time of code generation, and at run-time the function only moves the data, computing the index sequences and detecting when each cycle finishes. Therefore even though the function is completely generic, supporting any dimensions, the most complex part of the problem is tackled programmatically during code generation, resulting in a very simple run-time code.

@@

Given a set of sorted arrays, once the tree is initialized data is taken from it one vector at a time, producing the final sorted sequence. Each time a new vector is requested the tree is traversed following the smallest element among the node children, until one of the input arrays is reached. The data from the array is then loaded from memory and merged with the intermediate data, producing new updates moving through the tree until the root is reached.

The intermediate data from the tree is kept by \chipsort in a contiguous array, treated as a heap. Assuming 1-based indexing, each node ancestor can be simply found by the Euclidean division of its index by 2. This data is assumed to fit some level of cache memory during operation, as in the related work~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15}.

\section{Experiments}
\label{sec:experiments}
%
This section reports experiments with the \chipsort methods, as well as alternative ones. The term {\it method} can be understood as both the algorithms and their implementations in the Julia language. The specific methods tested are:
\begin{itemize}
\item {\tt chipsort!} Combsort finishing with Insertion sort, written as a serial program and vectorized by the compiler.
\item {\tt chipsort\_small!} in-register, based on sorting networks.
\item {\tt chipsort\_medium!} vectorized Combsort, plus additional steps, finishing with Insertion sort.
\item {\tt chipsort\_large} merge-sort with a multi-way merge tree, starting from chunks processed by {\tt chipsort\_medium!}.
\item {\tt insertion\_sort!} Insertion sort taken from the Julia library.
\item {\tt sort!} the Julia standard sort, Quicksort finishing with Insertion sort for small arrays (\href{https://github.com/JuliaLang/julia/blob/de3401908a506727ed70b3e953bd721ab4d66109/base/sort.jl#L461}{20} elements).
\end{itemize}

@@

\subsection{Medium and large arrays}
Our second experiment compared many different methods on inputs of exponentially increasing sizes, from $2^6$ to $2^{20}$. Figure~\ref{fig:bench-time} displays the resulting measurements. Although the curves look too concentrated in this visualization, it is possible to notice that {\tt chipsort!} consistently outperforms the Julia standard library by a factor of 80\% to 100\%, or up to twice as fast. We can also observe that Insertion sort soon becomes too inefficient.

\begin{figure}[htb]
\centerline{\includegraphics[width=\linewidth]{fig/chipsort-bench-time.pdf}}
\caption{Running time of the different sorting methods studied.}
\label{fig:bench-time}
\end{figure}

Figure~\ref{fig:bench-curves} shows a different visualization from the same data with the time divided by the input size. The vertical dotted lines indicate the size of each cache level. Here we can see more clearly how {\tt chipsort\_medium!} and {\tt chipsort\_large} slightly outperform {\tt chipsort!} in a few cases.

\begin{figure}[htb]
\centerline{\includegraphics[width=\linewidth]{fig/chipsort-bench-curves.pdf}}
\caption{Curves displaying the relative running time from each studied method. The ordinate represents the median of the measured running times divided by the input size at each test. Lower is faster.}
\label{fig:bench-curves}
\end{figure}

@@

\begin{figure*}[htb]
  \centerline{\includegraphics[width=0.4\linewidth]{fig/chipsort-bench-8k.pdf}
    \includegraphics[width=0.4\linewidth]{fig/chipsort-bench-256k.pdf}}
\caption{Empirical CDF from different methods with 8k and 256k {\tt UInt32} values. Results from 10,000 measurements.}
\label{fig:bench-cdf}
\end{figure*}

\subsection{Other languages}
A simple experiment was also carried out with different programming languages in the same machine, and with Julia in different machines to provide a richer context. Table~\ref{tab:bench-cxx} displays statistics collected from Julia, Python and C++ programs. For the case of Python the measurements obtained by calling the function through PyCall were better than using the native Python shell, therefore the best value was kept. Results obtained for C++ with Julia inter-operation using Cxx.jl were slightly worse than native, thus discarded. While Numpy was almost as fast as Julia, no reason could be found for the dismal performance of the C++ standard library.

\begin{table}[h]
\tabcolsep11pt
\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different software platforms.}{
\begin{tabular}{|l|r|r|}\hline
Method & 8k & 1M\\\hline
\chipsort & 172.35 $\mu$s & 35.28 ms\\\hline
Julia standard & 345.46 $\mu$s & 68.39 ms\\\hline
Numpy via PyCall & 369.50 $\mu$s & 70.01 ms\\\hline
C++ {\tt std::qsort} & 788.59 $\mu$s & 147.37 ms\\\hline
C++ {\tt std::sort} & 1,207.23 $\mu$s & 224.71 ms\\\hline
\end{tabular}}
\label{tab:bench-cxx}
\end{table}

Table~\ref{tab:bench-512} contains the results for {\tt chipsort!} and the Julia standard {\tt sort!} in an Intel processor with AVX512 instructions, and an ARM processor with NEON instructions.
\begin{table}[h]
\tabcolsep11pt
\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different hardware platforms.}{
\begin{tabular}{|l|l|r|r|}\hline
Chip & Method & 8k & 1M\\\hline
AXV512 & \chipsort & 177.4 $\mu$s & 35.20 ms\\\hline
AXV512 & Julia standard & 347.5 $\mu$s & 69.07 ms\\\hline
NEON & \chipsort & 350.9 $\mu$s & 73.36 ms\\\hline
NEON & Julia standard & 573.2 $\mu$s & 113.68 ms\\\hline
\end{tabular}}
\label{tab:bench-512}
\end{table}

No change in performance can be seen in the AVX512 machine, and on the ARM machine the speedup was smaller than before: only 60\%. The most notable fact in this experiment is that the same high-level and serial code successfully produced a vectorized sort on machines with different vector sizes and even instruction sets, always with a beneficial outcome.

\section{Conclusion}
\label{sec:conclusion}
%
We have presented \chipsort, a Julia module for SIMD and cache-aware sorting. The main sorting method utilized in \chipsort is Combsort finalized with Insertion sort. This method is based on previous proposals for SIMD sorting~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} and other practical sorting methods~\cite{INCERPI198737,musser1997introspective}. This method was implemented as a serial program vectorized by the Julia compiler and LLVM, resulting in a sorting function up to twice as fast as the Julia standard library sort. This performance holds for at least a million 32-bit integers in a personal computer.

@@

While the main \chipsort method relies on compiler optimizations, other methods utilize explicit vectorization trough SIMD.jl. \chipsort makes extensive use of metaprogramming, more specifically Julia generated functions. This allows functions like sorting networks to be implemented in an abstract and generic way. The networks are represented as data structures, used by a higher-level program to assemble the final network code. This contrasts with other libraries where different networks are implemented straight as final functions. Julia also has the opportunity to perform custom optimizations since \chipsort is implemented in pure Julia with abstract methods that can be specialized.

It is a testament to the quality of the Julia compiler infrastructure that the most successful method implemented in \chipsort has no explicit vectorization or intrinsics. It is just a simple, scalar implementation of Combsort that is optimized by the compiler to generate code adapted to different problem settings and to the hardware architecture at run-time. And in the cases where explicit vectorization is necessary Julia still offers a suitable framework that allowed \chipsort to implement sophisticated techniques with concise code. This project can illustrate how well Julia implements traditional language features such as parametric types, and its metaprogramming features make it stand out from other languages even more.

Some future prospects for \chipsort are to implement missing features such as stable sorting of key-payload data and buffers in the merge tree; explore the limits of in-register processing in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to optimize code for complex tasks such as multi-stage multi-threaded merge-sorting of large arrays.
